---
title: "MARSS and Hwange data"
date: "24/09/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dpi = 300)
```


## Read in and visualize data

Prerequisites.
```{r}
library(tidyverse)
theme_set(theme_light()) 
library(janitor)
library(lubridate)
library(MARSS)
```

Read in the data.
```{r}
wps7293 <- readxl::read_xls("data/Water point surveys 1972 - 1993 - Total Number.xls") %>% clean_names()
wps2001 <- readxl::read_xls("data/Water point surveys 1994 - 2001 - Detailled.xls") %>% clean_names()
wps2005 <- readxl::read_xls("data/Water point surveys 2002 - 2005 - Detailed.xls") %>% clean_names()
```

```{r}
wps7293 %>%
  count(water_point, sort = TRUE)

wps7293 %>%
  count(species, sort = TRUE)

wps7293 %>%
  count(year, sort = TRUE)
```



```{r}
dat <- wps7293 %>%
  filter(species %in% c("Elephant", "Giraffe", "Lion", "Impala")) %>%
  select(year, species, total) %>%
  mutate(species = as_factor(species))
dat
```

Plot all data.
```{r}
dat %>%
  group_by(year, species) %>%
  summarise(mean_biomass = round(mean(total))) %>%
  ggplot() + 
  aes(x = year, y = mean_biomass) + 
  geom_point() +
  geom_smooth() + 
  labs(x = "Year", 
       y = "Counts",
       color = "Species") +
  facet_wrap(~species, scales = "free")
```

## MARSS model in the frequentist framework

We consider a model with the following assumptions:
* All prey species share the same process variance.
* All predator species share the same process variance.
* Prey and predator species have different measurement variances.
* Measurement errors are independent.
* Process errors are independent.

We fit this model with the `MARSS` package. We need to specify the ingredients first. 
```{r}
Q <- matrix(list(0), 4, 4)
diag(Q) <- c("Prey", "Prey", "Prey", "Predator")
R <- matrix(list(0), 4, 4)
diag(R) <- c("Prey", "Prey", "Prey", "Predator")
model.0 <- list(
  B = "unconstrained", U = "zero", Q = Q,
  Z = "identity", A = "zero", R = R,
  x0 = "unequal", tinitx = 1
)
model.0
```
Then we fit the model.
```{r}
mod.0 <- dat %>%
  group_by(year, species) %>%
  summarise(mean_biomass = round(mean(total))) %>%
  ungroup() %>%
  pivot_wider(names_from = species, values_from = mean_biomass) %>%
  select(-year) %>%
  t() %>%
  MARSS(model = model.0)
```

We may get the estimates in a more readable format. For example, let's have a look to the interactions. These estimates describe the effect of the density of species $j$ on the per capita growth rate of species $i$.
```{r}
B.0 <- coef(mod.0, type = "matrix")$B[1:4, 1:4]
rownames(B.0) <- colnames(B.0) <- c("Elephant", "Giraffe", "Impala", "Lion")
print(B.0, digits = 2)
```

The effect of species $j$ on species $i$ is given by the cell at $i$-th row and $j$-th column. The B matrix suggests that Lion has a negative effect on impala and positive on Elephant and Giraffe. In the diagonal, we have the strength of density-dependence: if species $i$ is density-independent, then $B_{i,i}$ equals 1, like Impala; smaller $B_{i,i}$ means more density dependence, like Giraffe.

Compare observations to fitted values.
```{r}
fr <- forecast.marssMLE(mod.0, h=0)
plot(fr)
```


Forecast 10 years ahead.
```{r}
fr <- forecast.marssMLE(mod.0, h = 10)
plot(fr)
```


Try again, with a simpler model.
```{r}
Q <- matrix(list(0), 4, 4)
diag(Q) <- c("all", "all", "all", "all")
R <- matrix(list(0), 4, 4)
diag(R) <- c("all", "all", "all", "all")
model.0 <- list(
  B = "unconstrained", U = "zero", Q = Q,
  Z = "identity", A = "zero", R = R,
  x0 = "unequal", tinitx = 1
)
model.0
```
Then we fit the model.
```{r}
mod.0 <- dat %>%
  group_by(year, species) %>%
  summarise(mean_biomass = round(mean(total))) %>%
  ungroup() %>%
  pivot_wider(names_from = species, values_from = mean_biomass) %>%
  select(-year) %>%
  t() %>%
  MARSS(model = model.0)
```

We may get the estimates in a more readable format. For example, let's have a look to the interactions. These estimates describe the effect of the density of species $j$ on the per capita growth rate of species $i$.
```{r}
B.0 <- coef(mod.0, type = "matrix")$B[1:4, 1:4]
rownames(B.0) <- colnames(B.0) <- c("Elephant", "Giraffe", "Impala", "Lion")
print(B.0, digits = 2)
```

The effect of species $j$ on species $i$ is given by the cell at $i$-th row and $j$-th column. The B matrix suggests that Lion has a negative effect on impala and positive on Elephant and Giraffe. In the diagonal, we have the strength of density-dependence: if species $i$ is density-independent, then $B_{i,i}$ equals 1, like Impala; smaller $B_{i,i}$ means more density dependence, like Giraffe.

Compare observations to fitted values.
```{r}
fr <- forecast.marssMLE(mod.0, h=0)
plot(fr)
```


Forecast 10 years ahead.
```{r}
fr <- forecast.marssMLE(mod.0, h = 10)
plot(fr)
```
